Detecting skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights

### 2022-09-01 manuscript/manuscript.Rmd
updated the readme file.

---
title: "Detecting skill level and mental templates in Late Acheulean biface morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean bifaces, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of biface morphology, including knapper skill levels and mental templates. Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of biface morphology. The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training using the data from a 90-hour-long knapping skill acquisition experiment. We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program. \\par \\textbf{Keywords:} Late Acheulean; Biface production; Boxgrove; Experimental archaeology; Skill level; Mental template"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The morphological variability of Acheulean bifaces has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the homogeneity and longevity of biface assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011]. More specifically, a complex suite of interconnecting factors have been identified to contribute to biface morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021]. From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of biface studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles. Of particular attention here are the experimental works conducted by Stout et al. [-@stout2014] focusing on inferring high knapping skill level and Garcia-Medrano et al. [-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage. Our paper combines these two perspectives and provides novel insights to the same archaeological assemblage by comparing it with experimentally made bifaces.

In its classical definition, the term mental template indicates that the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45]. This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework. Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applies it in the typological analysis of bifaces at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004]. Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986]. For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignorance of other competing factors such as raw material constraints [@white1995], and c) the lack of rigorous studies of its corresponding cognitive processes. To avoid the historical baggage associated with this controversial term, some researchers developed alternative frameworks such as "design imperatives" purely derived from ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].

Until recently, researchers have actively addressed the above-mentioned critiques and reconceptualized the mental template in the study of biface morphology. Regarding the normative and static assumptions, Hutchence and Scott [-@hutchence2021], for example, leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations, especially how the social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning. They further emphasized that emergent actions of individual knappers also contribute greatly to the shape of Boxgrove handaxes but they were simultaneously constrained by the imposition of social norms. This view also somewhat echoes the "individualized memic construct" proposed by McNabb et al. [-@mcnabb2004], which tries to provide a more balanced perspective incorporating both individual agency and social learning. As for the critique towards confounding factors explaining morphological variability, raw material is often treated as an important variable to be controlled at the very beginning of a research design focusing on mental templates. This is best exemplified by an experimental study of García-Medrano et al. [-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in the Boxgrove archaeological assemblage in composition, size, and shape. In terms of the cognitive mechanisms behind mental templates, Ho and colleagues [-@ho2022] recently developed a series of navigation experiments demonstrating the externalization of the planning process to simple geometric representations instead of a complete representation of the given task, featuring both the efficiency and flexibility given the limited cognitive resources. Their experimental design has the potential to be transferred into a research setting aiming at directly testing the planning of knapping behaviors and elucidating how "mental templates" are constructed and perceived in brains. In short, when exercised with proper caution, the concept of mental templates still has its value in our study of biface morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.

Following the reconceptualization of the mental template as a more flexible and interactive concept, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66]. This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989]. It also roughly corresponds to the well-known dichotomy developed by French lithic analysts of "*connaissance*" (abstract knowledge) and "*savoir-faire*" (practical know-how) [@pelegrin1993]. As Stout [-@stout2002: 694] noted, the acquisition of skill is deeply rooted in its social context, and it is not composed of "some rigid motor formula" but "how to act in order to solve a problem". This ecological notion of skill somewhat mirrors Hutchence and Scott's [-@hutchence2021] reconceptualization of the mental template in that they both refute the idea that technology is simply an internal program expressed by the mind and they prefer a dynamic approach emphasizing the interaction between perception and action. The manifestations of skill in materialized form display a great amount of variation, but ethnoarchaeological studies have repeatedly suggested that skills can be improved through practice as perceived by the local practitioners. It is thus possible to evaluate the skill levels reflected in knapping products [@stout2002; @roux1995]. When contextual information is less readily available as in the Late Acheulean archaeological assemblages, how to properly operationalize and measure knapping skills has been a methodological issue receiving much attention among archaeologists [@bamforth2008; @kolhatkar2022]. In addition to measurements that can be almost applied in any lithic technological system such as raw materials, platform preparation, as well as hinges, in the context of biface technology, symmetry [@hodgson2015; @hutchence2019] and cross-sectional thinning [@caruana2020; @stout2014; @pargeter2019] have been frequently quoted as reliable and distinctive indicators of the skill level as supported by several experimental studies. These two features have also been commonly used as standards for dividing Early Acheulean and Late Acheulean [@schick1993; @clark2001; @callahan1979].

Drawing on these two lines of literature, we aim to explore the possibility of differentiating skill level and mental template and the interaction between the two through a comparative study of an archaeological biface assemblage known for its remarkable dexterity, a reference biface collection produced by modern knapping experts, and an experimental biface sample produced by modern novice knappers. We generated the novice biface collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest. As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental templates be efficiently detected from biface morphometric data? 2) How does training affect novices' performance in these two aspects?

@@

## Boxgrove biface collection

@@

## Experimental biface collection

The biface experimental replicas used in this study comprised two sub-collection. The first sub-collection includes 10 bifaces knapped by three expert knappers, including Bruce Bradley (n=4), John Lord (n=3), and Dietrich Stout (n=3) [@stout2014]. These handaxes were made for previous research projects, which similarly aimed to approximate 'Late Acheulean' handaxes explicitly comparable to the Boxgrove assemblage [@faisal2010; @stout2011a; @stout2014]. The second sub-collection is produced from a 90-hour handaxe knapping skill acquisition experiment [@bayani2021; @pargeter2019; @pargeter2020], where 30 adults with no previous experience in knapping were recruited from Emory University and its surrounding communities and requested to make 132 bifaces in total. Among these 30 adult participants, 17 have gone through multiple one-to-one or group training sessions that amounted to 89 hours in maximum, while the remaining 13 were assigned to the controlled group, where no formal training is given. As part of the preparation efforts, the experimental team spalled the Norfolk flints acquired through [Neolithics.com](https://neolithics.com/) into flat blanks of similar size and shape for training and assessments. The mechanical properties of these raw materials are comparable to the ones used in Boxgrove in that they are both fine-grained and highly predictable in fracturing process.

@@

To better understand the morphological variation of Boxgrove biface collection, we adopted a standardized analytical procedure to extract the morphometric information from 752 photos of the studied samples [@iovita2011], which include both the front and lateral views of a given specimen. First, we used Adobe Photoshop to conduct a batch transformation of the samples' pixel scale into a real-world measurement scale based on the fixed photographic setting. This is then followed by the batch conversion of color photographs to a black-and-white binary format. Subsequently, we cropped the silhouettes of bifaces one by one using the Quick Selection Tool in Adobe Photoshop. The metric measurements were conducted in ImageJ [@rueden2017], where we employed a custom ImageJ script [@pargeter2019] to measure the maximum length, width, and thickness of a given silhouette. The width and thickness measurements are taken at 10% increments of length starting at the tip of each biface (**Figure** \@ref(fig:ImageJ)), which eventually leads to 19 morphometric variables in total (1 length measurement, 9 width measurements, and 9 thickness measurements). Finally, we calculated the geometric means of all 19 linear measurements to create a scale-free data set that preserves the individual morphological variation at the same time [@lycett2006]. This allometric scaling procedure controls for size variation which may come from initial blanks and/or reduction intensity (shaping/resharpening). Notably, Shipton and Clarkson [-@shipton2015] previously found that reduction intensity does not have a strong impact on the shape of handaxes. The same procedure was also applied to the morphometric analyses of the experimental biface collection, which was partially published in Pargeter et al. [-@pargeter2019].

```{r ImageJ, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A visual demonstration of the biface measurement protocol using Image J (after Pargeter et al. 2019: Figure 5)."}
knitr::include_graphics(here::here("figure/Fig1.png"))
```

@@

Given the number of variables involved in this study, we used principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020a; @garcía-medrano2020b; @herzlinger2017]. To detect the effect of training on novices' performance as compared with archaeological samples and biface made by experts, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test, which does not rely on the assumptions of equal sample sizes and equal variance. This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which can be accessed through the author's Github (<https://github.com/Raylc/PaST-pilot>).

@@

Our analysis suggested that the first two components already explain 77.2% of the variation for the entire morphometric data set composed of 19 variables (**Figure** \@ref(fig:Screeplot)), which is a rather reasonable variance ratio to avoid overfitting. Variable loadings (**Table** \@ref(tab:tab1)) indicate that the first principal component (PC1) captures overall cross-sectional thickness. It is positively correlated with all thickness measurements while negatively correlated with all other measurements. A higher PC1 value thus indicates a thicker biface, and vice versa. The second principal component (PC2) tracks elongation and pointedness, as indicated by a positive covariance of maximum length and bottom width/thickness. As PC2 increases, a biface will be relatively longer and more convergent from the broad base to the tip.. Thus, PC1 corresponds to cross-sectional thinning and PC2 to overall shape variation.

@@

A closer look at the principal component scatter plot (**Figure** \@ref(fig:GeneralPCA1)) yields the clustering of different groups of bifaces. The majority of Boxgrove bifaces occupy an area featuring negative values of both PC1 and PC2. The expert group is similar to the Boxgrove group in PC1, while the former has a relatively higher PC2 value than the latter on average. The group of novice displays the highest level of variability, however, it is rather pronounced that most bifaces made by novices have a positive PC1 value that is different from both the groups of Boxgrove and experts.

```{r GeneralPCA1, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A principal component scatter plot of bifaces from the groups of Boxgrove (red, n=326), expert (green, n=10), and novice (blue, n=132)."}
knitr::include_graphics(here::here("figure/Fig3.png"))
```

@@

We extracted the PC1 and PC2 values of individual bifaces and compared them between different groups. More specifically, the novice group was divided into three sub-groups based on their training stages as specified in the method section. As such, we found that for PC1 values (**Figure** \@ref(fig:PCA1)), the only two group comparisons that are **not** statistically significant are the one between Boxgrove and Expert ($t=-1.65$, $p>0.05$) and the one between Early training and Late training stages ($t=-0.649$, $p>0.05$), which at least partially confirms our visual observation of the general PCA scatter plot. Likewise, for PC2 values (**Figure** \@ref(fig:PCA2)), the group comparison between the Early training and Late stages again is not statistically significant ($t=0.333$, $p>0.05$). An unexpected result is that the mean PC2 value difference between the Pre-training group and Boxgrove is also not statistically significant ($t=-0.818$, $p>0.05$).

@@

Our study suggested that both skill level and mental template have a relatively clear manifestation in different aspects of biface morphology, where the former is related to cross-sectional thinning (PC1) while the latter relates to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program. In accordance with the existing literature on biface knapping skill [@caruana2020; @stout2014; @callahan1979], the results of PCA suggested that PC1 (cross-sectional thinning) is a robust indicator of skill level as it is a common feature shared by modern expert knapper and Boxgrove knappers. Thinning is regarded as a technique requiring a high knapping skill level because it requires one to carefully detach flakes in an invasive manner while not breaking the biface into several pieces, serving the purpose of achieving the desired convexity and/or volume. This procedure involves precise control of striking forces, strategic choice of platform external angle, and attentive preparation of bifacial intersection plane, all of which were part of our experimental training program [@callahan1979; @caruana2022; @stout2014; @pargeter2020; @shipton2013]. Experimental studies have also shown that the thinning stage of biface produce often involves the use of soft hammers, which is also supported by indirect archaeological evidence of flake attributes from Boxgrove [@roberts2009; @roberts1998: 384-394]. This also reflects the majority of samples in both our expert and novice experimental assemblages. In the skill acquisition experiments, novice knappers have been explicitly taught to switch to the soft hammer for thinning purposes, but some of them did not follow the instruction during the assessment. On the other hand, it has also been shown that hard hammers can also be used to achieve similar thinning results [@pelcin1997; @bradley1986], corresponding to the cases of replicas produced by Bruce Bradley.

>>>>

---
title: "Detecting skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates. Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of handaxe morphology. The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training using the data from a 90-hour-long knapping skill acquisition experiment. We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program. \\par \\textbf{Keywords:} Late Acheulean; Handaxe production; Boxgrove; Experimental archaeology; Skill level; Mental template"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The morphological variability of Acheulean handaxes has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011]. More specifically, a complex suite of interconnecting factors have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021]. From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of handaxe studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles. Of particular attention here are the experimental works conducted by Stout et al. [-@stout2014] focusing on inferring high knapping skill level and Garcia-Medrano et al. [-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage. Our paper combines these two perspectives and provides novel insights to the same archaeological assemblage by comparing it with experimentally made handaxes.

In its classical definition, the term mental template indicates that the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45]. This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework. Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applies it in the typological analysis of handaxes at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004]. Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986]. For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignorance of other competing factors such as raw material constraints [@white1995], and c) the lack of rigorous studies of its corresponding cognitive processes. To avoid the historical baggage associated with this controversial term, some researchers developed alternative frameworks such as "design imperatives" purely derived from ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].

Until recently, researchers have actively addressed the above-mentioned critiques and reconceptualized the mental template in the study of handaxe morphology. Regarding the normative and static assumptions, Hutchence and Scott [-@hutchence2021], for example, leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations, especially how the social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning. They further emphasized that emergent actions of individual knappers also contribute greatly to the shape of Boxgrove handaxes but they were simultaneously constrained by the imposition of social norms. This view also somewhat echoes the "individualized memic construct" proposed by McNabb et al. [-@mcnabb2004], which tries to provide a more balanced perspective incorporating both individual agency and social learning. As for the critique towards confounding factors explaining morphological variability, raw material is often treated as an important variable to be controlled at the very beginning of a research design focusing on mental templates. This is best exemplified by an experimental study of García-Medrano et al. [-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in the Boxgrove archaeological assemblage in composition, size, and shape. In terms of the cognitive mechanisms behind mental templates, Ho and colleagues [-@ho2022] recently developed a series of navigation experiments demonstrating the externalization of the planning process to simple geometric representations instead of a complete representation of the given task, featuring both the efficiency and flexibility given the limited cognitive resources. Their experimental design has the potential to be transferred into a research setting aiming at directly testing the planning of knapping behaviors and elucidating how "mental templates" are constructed and perceived in brains. In short, when exercised with proper caution, the concept of mental templates still has its value in our study of handaxe morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.

Following the reconceptualization of the mental template as a more flexible and interactive concept, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66]. This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989]. It also roughly corresponds to the well-known dichotomy developed by French lithic analysts of "*connaissance*" (abstract knowledge) and "*savoir-faire*" (practical know-how) [@pelegrin1993]. As Stout [-@stout2002: 694] noted, the acquisition of skill is deeply rooted in its social context, and it is not composed of "some rigid motor formula" but "how to act in order to solve a problem". This ecological notion of skill somewhat mirrors Hutchence and Scott's [-@hutchence2021] reconceptualization of the mental template in that they both refute the idea that technology is simply an internal program expressed by the mind and they prefer a dynamic approach emphasizing the interaction between perception and action. The manifestations of skill in materialized form display a great amount of variation, but ethnoarchaeological studies have repeatedly suggested that skills can be improved through practice as perceived by the local practitioners. It is thus possible to evaluate the skill levels reflected in knapping products [@stout2002; @roux1995]. When contextual information is less readily available as in the Late Acheulean archaeological assemblages, how to properly operationalize and measure knapping skills has been a methodological issue receiving much attention among archaeologists [@bamforth2008; @kolhatkar2022]. In addition to measurements that can be almost applied in any lithic technological system such as raw materials, platform preparation, as well as hinges, in the context of handaxe technology, symmetry [@hodgson2015; @hutchence2019] and cross-sectional thinning [@caruana2020; @stout2014; @pargeter2019] have been frequently quoted as reliable and distinctive indicators of the skill level as supported by several experimental studies. These two features have also been commonly used as standards for dividing Early Acheulean and Late Acheulean [@schick1993; @clark2001; @callahan1979].

Drawing on these two lines of literature, we aim to explore the possibility of differentiating skill level and mental template and the interaction between the two through a comparative study of an archaeological handaxe assemblage known for its remarkable dexterity, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers. We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest. As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental templates be efficiently detected from handaxe morphometric data? 2) How does training affect novices' performance in these two aspects?

@@

## Boxgrove handaxe collection

@@

## Experimental handaxe collection

The handaxe experimental replicas used in this study comprised two sub-collection. The first sub-collection includes 10 handaxes knapped by three expert knappers, including Bruce Bradley (n=4), John Lord (n=3), and Dietrich Stout (n=3) [@stout2014]. These handaxes were made for previous research projects, which similarly aimed to approximate 'Late Acheulean' handaxes explicitly comparable to the Boxgrove assemblage [@faisal2010; @stout2011a; @stout2014]. The second sub-collection is produced from a 90-hour handaxe knapping skill acquisition experiment [@bayani2021; @pargeter2019; @pargeter2020], where 30 adults with no previous experience in knapping were recruited from Emory University and its surrounding communities and requested to make 132 handaxes in total. Among these 30 adult participants, 17 have gone through multiple one-to-one or group training sessions that amounted to 89 hours in maximum, while the remaining 13 were assigned to the controlled group, where no formal training is given. As part of the preparation efforts, the experimental team spalled the Norfolk flints acquired through [Neolithics.com](https://neolithics.com/) into flat blanks of similar size and shape for training and assessments. The mechanical properties of these raw materials are comparable to the ones used in Boxgrove in that they are both fine-grained and highly predictable in fracturing process.

@@

To better understand the morphological variation of Boxgrove handaxe collection, we adopted a standardized analytical procedure to extract the morphometric information from 752 photos of the studied samples [@iovita2011], which include both the front and lateral views of a given specimen. First, we used Adobe Photoshop to conduct a batch transformation of the samples' pixel scale into a real-world measurement scale based on the fixed photographic setting. This is then followed by the batch conversion of color photographs to a black-and-white binary format. Subsequently, we cropped the silhouettes of handaxes one by one using the Quick Selection Tool in Adobe Photoshop. The metric measurements were conducted in ImageJ [@rueden2017], where we employed a custom ImageJ script [@pargeter2019] to measure the maximum length, width, and thickness of a given silhouette. The width and thickness measurements are taken at 10% increments of length starting at the tip of each handaxe (**Figure** \@ref(fig:ImageJ)), which eventually leads to 19 morphometric variables in total (1 length measurement, 9 width measurements, and 9 thickness measurements). Finally, we calculated the geometric means of all 19 linear measurements to create a scale-free data set that preserves the individual morphological variation at the same time [@lycett2006]. This allometric scaling procedure controls for size variation which may come from initial blanks and/or reduction intensity (shaping/resharpening). Notably, Shipton and Clarkson [-@shipton2015] previously found that reduction intensity does not have a strong impact on the shape of handaxes. The same procedure was also applied to the morphometric analyses of the experimental handaxe collection, which was partially published in Pargeter et al. [-@pargeter2019].

```{r ImageJ, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A visual demonstration of the handaxe measurement protocol using Image J (after Pargeter et al. 2019: Figure 5)."}
knitr::include_graphics(here::here("figure/Fig1.png"))
```

@@

Given the number of variables involved in this study, we used principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020a; @garcía-medrano2020b; @herzlinger2017]. To detect the effect of training on novices' performance as compared with archaeological samples and handaxe made by experts, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test, which does not rely on the assumptions of equal sample sizes and equal variance. This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which can be accessed through the author's Github (<https://github.com/Raylc/PaST-pilot>).

@@

Our analysis suggested that the first two components already explain 77.2% of the variation for the entire morphometric data set composed of 19 variables (**Figure** \@ref(fig:Screeplot)), which is a rather reasonable variance ratio to avoid overfitting. Variable loadings (**Table** \@ref(tab:tab1)) indicate that the first principal component (PC1) captures overall cross-sectional thickness. It is positively correlated with all thickness measurements while negatively correlated with all other measurements. A higher PC1 value thus indicates a thicker handaxe, and vice versa. The second principal component (PC2) tracks elongation and pointedness, as indicated by a positive covariance of maximum length and bottom width/thickness. As PC2 increases, a handaxe will be relatively longer and more convergent from the broad base to the tip.. Thus, PC1 corresponds to cross-sectional thinning and PC2 to overall shape variation.

@@

A closer look at the principal component scatter plot (**Figure** \@ref(fig:GeneralPCA1)) yields the clustering of different groups of handaxes. The majority of Boxgrove handaxes occupy an area featuring negative values of both PC1 and PC2. The expert group is similar to the Boxgrove group in PC1, while the former has a relatively higher PC2 value than the latter on average. The group of novice displays the highest level of variability, however, it is rather pronounced that most handaxes made by novices have a positive PC1 value that is different from both the groups of Boxgrove and experts.

```{r GeneralPCA1, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A principal component scatter plot of handaxes from the groups of Boxgrove (red, n=326), expert (green, n=10), and novice (blue, n=132)."}
knitr::include_graphics(here::here("figure/Fig3.png"))
```

@@

We extracted the PC1 and PC2 values of individual handaxes and compared them between different groups. More specifically, the novice group was divided into three sub-groups based on their training stages as specified in the method section. As such, we found that for PC1 values (**Figure** \@ref(fig:PCA1)), the only two group comparisons that are **not** statistically significant are the one between Boxgrove and Expert ($t=-1.65$, $p>0.05$) and the one between Early training and Late training stages ($t=-0.649$, $p>0.05$), which at least partially confirms our visual observation of the general PCA scatter plot. Likewise, for PC2 values (**Figure** \@ref(fig:PCA2)), the group comparison between the Early training and Late stages again is not statistically significant ($t=0.333$, $p>0.05$). An unexpected result is that the mean PC2 value difference between the Pre-training group and Boxgrove is also not statistically significant ($t=-0.818$, $p>0.05$).

@@

Our study suggested that both skill level and mental template have a relatively clear manifestation in different aspects of handaxe morphology, where the former is related to cross-sectional thinning (PC1) while the latter relates to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program. In accordance with the existing literature on handaxe knapping skill [@caruana2020; @stout2014; @callahan1979], the results of PCA suggested that PC1 (cross-sectional thinning) is a robust indicator of skill level as it is a common feature shared by modern expert knapper and Boxgrove knappers. Thinning is regarded as a technique requiring a high knapping skill level because it requires one to carefully detach flakes in an invasive manner while not breaking the handaxe into several pieces, serving the purpose of achieving the desired convexity and/or volume. This procedure involves precise control of striking forces, strategic choice of platform external angle, and attentive preparation of bifacial intersection plane, all of which were part of our experimental training program [@callahan1979; @caruana2022; @stout2014; @pargeter2020; @shipton2013]. Experimental studies have also shown that the thinning stage of handaxe produce often involves the use of soft hammers, which is also supported by indirect archaeological evidence of flake attributes from Boxgrove [@roberts2009; @roberts1998: 384-394]. This also reflects the majority of samples in both our expert and novice experimental assemblages. In the skill acquisition experiments, novice knappers have been explicitly taught to switch to the soft hammer for thinning purposes, but some of them did not follow the instruction during the assessment. On the other hand, it has also been shown that hard hammers can also be used to achieve similar thinning results [@pelcin1997; @bradley1986], corresponding to the cases of replicas produced by Bruce Bradley.

Given the dissimilarity of PC2 (elongation and pointedness) values between archaeological and experimental samples and its similarity among modern knappers, we argue that this dimension reflects different mental templates, where the Boxgrove assemblage displays an ovate shape featuring a wider tip while the experimental assemblages are characterized by a more pointed shape with a longer central axis. Our results regarding the ovate plan morphology of the Boxgrove assemblage generally supports what have been reported by Shipton and White [-@shipton2020] as well as Garcia-Medrano et al. [-@garcía-medrano2019]. This pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998] as advocated by Hutchence and Scott in handaxe analysis[-@hutchence2021]. The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise through actively teaching multiple novices at the same time. Meanwhile, novices had the chance to also communicate and learn from their peers, producing a shared repertoire of artifacts and actions. Unfortunately, the handaxe data from the instructor (N. Khreisheh) are unavailable, but it should be noted that the instructor has learned how to knap and how to teach knapping from one of our expert knapper (Bruce Bradley). This cascading effect of social learning might explain why there is a shared mental template between the expert group and the novice group after training.
